A Mechanical Formalization of Higher-Ranked Polymorphic Type Inference

### 2019-02-20 sections/Overview.tex
some small fixed in text after merge
Sections 2.3 and 2.4

Many rules are standard.
Rule $\mathtt{DeclVar}$ looks up term variables in the context.
Rules $\mathtt{Decl1I}$ and $\mathtt{Decl1I{\To}}$ respectively check and synthesize the unit type.
Rule $\mathtt{DeclAnno}$ synthesizes the annotated type $A$ of the annotated expression $(e:A)$
and checks that $e$ has type $A$.
Checking an expression $e$ against a polymorphic type $\all A$ in the context $\Psi$ succeeds
if $e$ checks against $A$ in the extended context $(\Psi, a)$.

The subsumption rule $\mathtt{DeclSub}$ depends on the subtyping relation,
and changes mode from checking to synthesis: if $e$ synthesizes type $A$ and $A\le B$
($A$ is more polymorphic than $B$), then $e$ checks against $B$.
If a checking problem does not match any other rules,
this rule can be applied to synthesize a type instead and then
check whether the synthesized type entails the checked type.

The Lambda abstraction is the most complicated construct of the bidirectional
type system. 
Checking $\lam e$ against function type $A\to B$ is easy:
we check the body $e$ against $B$ in the context extended with $x:A$.
However, synthesizing a lambda-function is a lot less easy, and 
this type system only synthesizes monotypes $\sigma\to\tau$.

@@

To avoid the problem of incrementally applying a substitution to remaining
tasks, we can find inspiration in the formulation of constraint solving
algorithms. For instance, the well-known unification
algorithm by \citet{unification} decomposes the problem of unifying two terms $s \stackrel{.}{=} t$ into a set
of related unification problems between pairs of terms $s_i \stackrel{.}{=} t_i)$. These smaller
problems are not tackled independently, but kept together in a set $S$. 
The algorithm itself is typically formulated as a small-step-style state
transition system $S \rightarrowtail S'$ that rewrites the set of unification
problems until it is in solved form or until a contradiction has been found.
For instance, the variable elimination rule is written as:
\[
   x \stackrel{.}{=} t, S  ~~\rightarrowtail~~  x \stackrel{.}{=} t, [t/x]S   \qquad\qquad{if}~x \not\in t~\text{and}~{x \in S}
\]
Because the whole set is explicitly available, the variable $x$ can be
simultaneously substituted everywhere.

@@

In recent work~\cite{itp2018}, we have shown how to mechanise a variant of DK's
subtyping algorithm and shown it correct with respect to DK's declarative
subtyping judgment. This approach overcomes the problem's in DK's formulation
by using a \emph{worklist}-based judgement of the form $$\Gamma \vdash \Omega$$
\jimmy{emphasize the form of judgment}
where $\Omega$ is a worklist, i.e. a sequence, of subtyping problems of the
form $A \leq B$.  The judgement is defined by case analysis on the first
element of $\Omega$ and recursively processes the worklist until it is empty.
Using both a single common ordered context $\Gamma$ and a worklist $\Omega$ greatly
simplifies propagating the instantiation of type variables that happen in one
subtyping problem to the remaining ones.

>>>>

Many rules are standard.
Rule $\mathtt{DeclVar}$ looks up term variables in the context.
Rules $\mathtt{Decl1I}$ and $\mathtt{Decl1I{\To}}$ respectively check and synthesize the unit type.
Rule $\mathtt{DeclAnno}$ synthesizes the annotated type $A$ of the annotated expression $(e:A)$
and checks that $e$ has type $A$.
Checking an expression $e$ against a polymorphic type $\all A$ in the context $\Psi$ succeeds
if $e$ checks against $A$ in the extended context $(\Psi, a)$.
The subsumption rule $\mathtt{DeclSub}$ depends on the subtyping relation,
and changes mode from checking to synthesis: if $e$ synthesizes type $A$ and $A\le B$
($A$ is more polymorphic than $B$), then $e$ checks against $B$.
If a checking problem does not match any other rules,
this rule can be applied to synthesize a type instead and then
check whether the synthesized type entails the checked type.

Lambda abstractions are the hardest construct of the bidirectional
type system to deal with. 
Checking $\lam e$ against function type $A\to B$ is easy:
we check the body $e$ against $B$ in the context extended with $x:A$.
However, synthesizing a lambda-function is a lot less easy, and 
this type system only synthesizes monotypes $\sigma\to\tau$.

@@

To avoid the problem of incrementally applying a substitution to remaining
tasks, we can find inspiration in the formulation of constraint solving
algorithms. For instance, the well-known unification
algorithm by \citet{unification} decomposes the problem of unifying two terms $s \stackrel{.}{=} t$ into a set
of related unification problems between pairs of terms $s_i \stackrel{.}{=} t_i$. These smaller
problems are not tackled independently, but kept together in a set $S$. 
The algorithm itself is typically formulated as a small-step-style state
transition system $S \rightarrowtail S'$ that rewrites the set of unification
problems until it is in solved form or until a contradiction has been found.
For instance, the variable elimination rule is written as:
\[
   x \stackrel{.}{=} t, S  ~~\rightarrowtail~~  x \stackrel{.}{=} t, [t/x]S   \qquad\qquad{if}~x \not\in t~\text{and}~{x \in S}
\]
Because the whole set is explicitly available, the variable $x$ can be
simultaneously substituted everywhere.

@@

In recent work, \citep{itp2018} have shown how to mechanise a variant of DK's
subtyping algorithm and shown it correct with respect to DK's declarative
subtyping judgment. This approach overcomes the problem's in DK's formulation
by using a \emph{worklist}-based judgement of the form $$\Gamma \vdash \Omega$$
where $\Omega$ is a worklist, i.e. a sequence, of subtyping problems of the
form $A \leq B$.  The judgement is defined by case analysis on the first
element of $\Omega$ and recursively processes the worklist until it is empty.
Using both a single common ordered context $\Gamma$ and a worklist $\Omega$ greatly
simplifies propagating the instantiation of type variables that happen in one
subtyping problem to the remaining ones.
